---
title: "Introduction"
author: "Cristian E. Nuno"
date: "`r format( x = Sys.Date(), format = '%B %d, %Y')`"
output: 
  github_document:
    dev: png
---

```{r setup, message = FALSE}
# load necessary packages ----
library(ISLR)       # data for ISLR examples
library(scales)     # scale functions for visualizations
library(tidyverse)  # data science packages

# load necessary objects ----
caption.text <- "Source: ISLR, 2017"
my.theme <- theme_minimal() + theme(panel.grid = element_blank())

# set dpi for all chunks ----
knitr::opts_chunk$set(dpi = 300)
```

# Introduction

* Statistical learning = understanding data
    + Supervised: Building a statistical model for predicting, or estimating, an output based on one or more inputs
    + Unsupervised: There are inputs but no supervising output

## Wage Data

In particular, we wish to understand the association between an employee’s age and education, as well as the calendar year, on his wage.

The [`Wage`](https://rdrr.io/cran/ISLR/man/Wage.html) data set involves predicting a **continuous** or **quantitative** output value, which is referred to as a **regression** problem.

```{r wage plots, message = FALSE}
# hourly wage described by age ----
Wage %>%
  ggplot(aes(x = age, y = wage)) +
  geom_point(color = "gray") +
  geom_smooth(se = FALSE, color = "navy") +
  xlab("Age") +
  ylab("Hourly wage") +
  scale_y_continuous(labels = dollar) +
  labs(title = "Whiles wages typically rise with age, they fall after age 60"
       , caption = caption.text) +
  my.theme

# hourly wage described by year ----
Wage %>%
  ggplot(aes(x = year, y = wage)) +
  geom_point(color = "gray") +
  geom_smooth(se = FALSE, method = "loess", color = "navy") +
  xlab("Year") +
  ylab("Hourly wage") +
  scale_y_continuous(labels = dollar) +
  labs(title = "On average, hourly wage rises slighly over time"
       , caption = caption.text) +
  my.theme

# hourly wage described by education ----
Wage %>%
  ggplot(aes(x = education, y = wage, fill = education)) +
  geom_boxplot() +
  scale_fill_brewer(name = "Education level", type = "qual", palette = "Accent") +
  xlab("Education level") +
  ylab("Hourly wage") +
  scale_y_continuous(labels = dollar) +
  labs(title = "On average, hourly wage rises with higher levels of educational attainment"
       , caption = caption.text) +
  my.theme
```

